Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-05-05 doc/paper.Rmd
First pass of revisions

institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights

@@

Highlights: `r rmarkdown::metadata$highlights`

@@

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
Although the local extinction of species from individual fragments can be common [@haddadHabitatFragmentationIts2015 @fletcher2018], the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009].
This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.

Climate change can alter the demographic of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015 @holyoak2016].
While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013 @duffy2015].
These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
These results, coupled with evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004, @selwood2015].

@@

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried (Figure S\_ <!--# map can go in the supplemental materials -->).
In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
While some *Heliconia* species are at least partially self-compatible [@kressSelfincompatibilitySystemsCentral1983], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare *[unpubl. data]*.

In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
The plots are located in continuous forest (n=6), 10 ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
This study uses data from 1-ha fragments and Continuous Forest sites.

Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants.
We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
Plants that were not found for three consecutive surveys were considered dead.
There is a seven-fold range in *H. acuminata* density in the forest plots, with density in several of the continuous forest plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].

@@

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmmTMPA3B432011]), though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
<!--# maybe add a sentence emphasizing that SPEI is monthly calculation -->

@@

<!--# I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).  -->

@@

The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
The general form of the vital rate models was as follows:

@@

where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see chapter 5 of @woodGeneralizedAdditiveModels2017 and @gasparriniPenalizedFrameworkDistributed2017 for more detail).
Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e. number of months) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex <!--# You may want to say what the 3 knots can capture: increasing then decreasing, exp decline etc -->.
Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].

@@

The `dlnm` package does not currently allow modeling of factor by smooth interactions.
This means we cannot \_\_\_\_\_\_\_\_\_\_\_\_\_\_.
We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.

@@

Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature.
For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).
<!--#  I know I'm missing some literature here. Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->

@@

Our dataset comprised `r summary_stats$n[2]` plants in continuous forest and `r summary_stats$n[1]` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` ).
When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
The proportion of surviving plants was lowest in the 2003-2004 transition year (P(surv) = 0.92), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
The lowest survival for 1-ha fragment plots (P(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006.
Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average \~`r (1 - (2.5/2.9))*100 %>% round()`% fewer shoots (2.5 ± 1.5 SD) and were \~`r (1 - (36.2/40.6))*100 %>% round()`% shorter (36.3 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently substantially larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

<!--# Is growth really higher in fragments?  Probably, since fig2 shows size becoming more similar over time.   -->

While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# not sure I can/should fill in that blank -->.
Even though flowering probability of large plants [log(size) \> 5] was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs. only 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
<!--# not sure how necessary this is. EB: same-->

<!-- This paragraphis talkng about size, but not really growth -->

@@

Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
For 1-ha fragments, recent SPEI values (i.e. within the past 0--12 months) had the strongest effect on survival, with the highest survival near SPEI of 0 and decreasing survival when plants were exposed to either drier or wetter conditions over the prior 0--12 months (Figure \@ref(fig:surv-spei)a).
In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 leading to slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year (i.e. not the most recent dry season, but the one before that), also affected survival probabilities in continuous forest.
Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
However, it should be noted that only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past.

Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-diff)).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.

Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 13.1 for continuous forest; see also Figure \@ref(fig:growth-spei)).
Drought at lags 8--11 (March--June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.

The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-diff).
However, growth in continuous forests is higher by up to log(size) = 0.57 when the current wet season is unusually wet (SPEI\>2).
This is due to reduced growth in fragments under wet conditions.

>>>>

institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; demographic vital rates; *Heliconia acuminata*; drought; ENSO; delayed effects; distributed lag non-linear models
highlights: |
  These are the highlights

@@

<!-- Highlights: `r rmarkdown::metadata$highlights` -->

@@

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015]. Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat. Although species in forest fragments can have depressed rates of reproduction or survivorship [\@zartmanDroughtImpactsSurvivorship2015; \@brunaSeedGerminationRainforest1999; \@lauranceRainForestFragmentation1998], leading to local or regional population declines, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]. However, it is often assumed that these detrimental effects are caused by dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]. The demographic consequences of habitat fragmentation are thought to be particularly important in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates. For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
Despite this, evidence linking changes in population-level demographic responses with changes in environmental conditions remains scarce.

Climate change has been shown to alter the demography of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015; @holyoak2016]. While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021]. Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015]. These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable. Plant populations may be particularly sensitive as an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019]. These results, coupled with evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004; @selwood2015].

@@

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried [@bierregaardBiologicalDynamicsTropical1992]. In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others, such as *H. acuminata,* are found at lower densities in the darker and cooler forest understory. These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991]. While some *Heliconia* species are at least partially self-compatible [@kressSelfincompatibilitySystemsCentral1983], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare *[unpubl. data]*.

In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004]. Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}=2$), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].

@@

In 1997--1998 we established a series of 5000 m^2^ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
The plots are located in continuous forest ($n=6$), 10 ha fragments ($n=3$), and 1-ha fragments ($n=4$), with distance between plots ranging from 500 m--41 km.
Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
This study uses data from 1-ha fragments and continuous forest sites.

Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering.
We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]. Plants that were not found for three consecutive surveys were considered dead. There is a seven-fold range in *H. acuminata* density in the forest plots, with density in several of the continuous forest plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].

@@

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º $\times$ 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993]. SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspirtaion of the previous three months.
SPEI calculations were made using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
<!--# add this as a "child document" or whatever -->

@@

@@

The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
The general form of the vital rate models was as follows: <!--# add flowering covariate? -->

@@

where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see chapter 5 of @woodGeneralizedAdditiveModels2017 and @gasparriniPenalizedFrameworkDistributed2017 for more detail). Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e. number of lagged SPEI values for each observation) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex. Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
Estimated degrees of freedom (edf) represent the 'true' complexity of the smooth after penalization with edf = 1 being equivalent to a straight line and larger numbers representing more complex curves.

@@

The `dlnm` package does not currently allow modeling of factor by smooth interactions.
This means we could not include habitat as an interaction term.
We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.

@@

Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature.
For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d). The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).

@@

Our dataset comprised `r format(summary_stats$n[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n[1], big.mark = ",")` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` SD; 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` SD).
When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
The proportion of surviving plants was lowest in the 2003--2004 transition year ($P\_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006.
Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on suvival in year t+1).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average \~`r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were \~`r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently substantially larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage.
Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# This requires a supplemental figure, probably.  A plot of shoots vs. log(size) -->.
Even though flowering probability of large plants [log(size) \> 5] was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs. only 12% in 1-ha fragments (Figure \@ref(fig:eda)c).

@@

Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
For 1-ha fragments, recent SPEI values (i.e. within the past 0--12 months) had the strongest effect on survival, with the highest survival near SPEI of 0 and decreasing survival when plants were exposed to either drier or wetter conditions over the prior 0--12 months (Figure \@ref(fig:surv-spei)a).
In continuous forest, recent SPEI had a weaker effect with only SPEI \< -1 leading to slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year (i.e. not the most recent dry season, but the one before that), also affected survival probabilities in continuous forest.
Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
However, it should be noted that only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past.
Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-diff)).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.

Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 13.1 for continuous forest; see also Figure \@ref(fig:growth-spei)).
Drought at lags 8--11 (March--June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-diff).
However, growth in continuous forests is higher by up to log(size) = 0.57 when the current wet season is unusually wet (SPEI\>2).
This is due to reduced growth in fragments under wet conditions.
